Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-03 intro.tex
+

At its core, the virtual memory functionality of modern CPUs is about \emph{location virtualization}: the memory locations (addresses) seen by most code are not, in fact, the exact location in physical memory where data resides. Instead these are \emph{virtual} addresses, which are mapped to actual physical resources by the cooperation of the hardware and OS. This is what enables separation of process memory resources:
the OS manipulates hardware functionality to ensure that any attempt by a process to access memory not explicitly granted to it by the kernel will fail.
But this is complicated by the fact that the OS and hardware can also enable, shared (overlapping) access to physical memory regions; the fact that the kernel data structures themselves are accessed via virtual memory addresses; and the fact that control over these mappings of virtual to physical addresses is itself mediated by \emph{in-memory data structures}, which the kernel still accesses via virtual address, leading to indirect cycles.

@@

This kind of context-dependent assertion, where a fact may be true in one address space but not others, has a modal flavor. 
We propose tackling the verification of virtual memory subsystems (and kernels more broadly) by adapting ideas from hybrid modal logic, which can label assertions true under \emph{other} circumstances (i.e., in another address space) with a modality indexed by a name for that space (in our case, the root of the page tables for an address space). This offers a convenient and powerful way to modularly (1) isolate assertions specific to a particular address space, (2) explicitly state when an assertion is true across address spaces, (3) manipulate address spaces from within other address spaces, and (4) reason about change in address spaces. These advantages make this approach to reasoning about virtual memory more flexible than prior program logic techniques~\cite{kolanski08vstte,kolanski09tphols}, which was only able to work with a single address space (the current address space on the CPU).

\begin{figure}
   \includegraphics[width=0.5\columnwidth]{architecture.png}
  \caption{Layers of Implementation}
  \label{fig:architecture}
  \end{figure}
We develop these ideas in an Iris-based\mytodo{cite JFP} separation logic, and verify simplified versions of several critical virtual-memory-related pieces of OS functionality, including mapping and unmapping pages, and describing the self-referential identity maps used in most kernels to allow convenient access to physical memory.

>>>>

At its core, the virtual memory functionality of modern CPUs is about \emph{location virtualization}: the memory locations (addresses) seen by most code are not, in fact, the exact location in physical memory where data reside. Instead these are \emph{virtual} addresses, which are mapped to actual physical resources by the cooperation of the hardware and OS. This is what enables separation of process memory resources:
the OS manipulates hardware functionality to ensure that any attempt by a process to access memory not explicitly granted to it by the kernel will fail.
But this is complicated by the fact that the OS and hardware can also enable, shared (overlapping) access to physical memory regions; the fact that the kernel data structures themselves are accessed via virtual memory addresses; and the fact that control over these mappings of virtual to physical addresses is itself mediated by \emph{in-memory data structures}, which the kernel still accesses via virtual address, leading to indirect cycles.

@@

This kind of context-dependent assertion, where a fact may be true in one address space but not others, has a modal flavor. 
We propose tackling the verification of virtual memory subsystems (and kernels more broadly) by adapting ideas from hybrid modal logic, which can label assertions true under \emph{other} circumstances (i.e., in another address space) with a modality indexed by a name for that space (in our case, the root of the page tables for an address space). This offers a \textit{convenient} and \textit{powerful way to modularly}
\begin{itemize}
\item isolate assertions specific to a particular address space,
\item explicitly state when an assertion is true across address spaces,
\item manipulate address spaces from within other address spaces, and
\item reason about change in address spaces.
\end{itemize}
  These advantages make this approach to reasoning about virtual memory more flexible than prior program logic techniques~\cite{kolanski08vstte,kolanski09tphols}, which was only able to work with a single address space (the current address space on the CPU).

\begin{figure}
   \includegraphics[width=0.5\columnwidth]{architecture.png}
  \caption{Layers of Implementation}
  \label{fig:architecture}
\end{figure}
We realize and show our reasoning principles, showin in Figure \ref{fig:architecture}
\begin{itemize}
\item we develop these ideas in an Iris-based\mytodo{cite JFP} separation logic, a logic for address-virtualization (\textsf{vProp} in Figure \ref{fig:architecture})
\item we model \textsf{AMD64} in Coq, and specify and prove its instructions in \textsf{vProp}
\item and verify simplified versions of several critical virtual-memory-related pieces of OS functionality, including mapping and unmapping pages, and describing the self-referential identity maps used in most kernels to allow convenient access to physical memory.
\end{itemize}
